Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph

### 2021-05-21 sdss-talk/slides.Rmd
updated sdss presentation

---
title: Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph
subtitle: SDSS 2021 Lightning Presentation
author: John Koo
date: 'June 2021'
output: 
  beamer_presentation:
    fig_crop: no
    theme: 'default'
    colortheme: 'beaver'
    includes:
      in_header: page_headers.tex
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \setbeamertemplate{itemize items}[circle]
---

@@

@@

By Screenshot taken by User:DarwinPeacock - Screenshot of free software GUESS, CC BY 3.0, https://commons.wikimedia.org/w/index.php?curid=6057981

@@

@@

# Generalized Ranodm Dot Product Graphs

@@

Want to estimate $X$ given $A$, or alternatively, 
interpoint distances, inner products, or angles.

@@

$$\max_i \|\hat{X}_i - Q_n X_i \| \stackrel{a.s.}{\to} 0
\text{ (Rubin-Delanchy et al., 2020)}$$

@@

@@

$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } A_{\cdot, i} = A c_i
\text{ and } c_i^{(i)} = 0$$

@@

$$\arg\min_{c_i} \|c_i\|_1 \text{ subject to } 
\hat{x}_i = \hat{X} c_i \text{ and } c_i^{(i)} = 0$$
$$A \approx \hat{X} I_{p, q} \hat{X}^\top$$

@@

**Theorem** (KTT): 
If $P = V \Lambda V^\top$ and $B = n V V^\top$, 
then $B_{ij} = 0$ $\forall i, j$ in different communities.

@@

**Theorem** (KTT): 
If $P = V \Lambda V^\top$ and $B = n V V^\top$, 
then $B_{ij} = 0$ $\forall i, j$ in different communities.

Orthogonal Spectral Clustering algorithm:

1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
positive and $K (K-1) / 2$ most negative eigenvalues.
2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
3. Construct graph $G$ using $B$ as its similarity matrix.
4. Partition $G$ into $K$ disconnected subgraphs.

**Theorem** (KTT): 
Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC. 
Then $\forall$ pairs $(i, j)$ belonging to different communities 
and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,

$$
\max_{i, j} \hat{B}^{(ij)} = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
$$

## Simulation Results

@@

IQR of community detection error rates using OSC (blue) compared against SSC on the ASE of A (purple), MM (red), and SSC on the adjacency matrix (green). Communities are approximately balanced. Simulations were repeated 50 times for each sample size.

>>>>

---
title: Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph
subtitle: SDSS 2021 Lightning Presentation
author: John Koo, Indiana University
date: 'June 2021'
output: 
  beamer_presentation:
    fig_crop: no
    theme: 'default'
    colortheme: 'beaver'
    includes:
      in_header: page_headers.tex
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \setbeamertemplate{itemize items}[circle]
---

@@

## Contributors

:::: {.column width='33%'}

```{r, out.width = '100px'}
knitr::include_graphics('john_koo.jpeg')
```

John Koo,  
PhD Student in Statistical Science,  
Indiana University

:::: {.column width='33%'}

```{r, out.width = '100px'}
knitr::include_graphics('minh_tang.jpg')
```

Minh Tang,  
Assistant Professor of Statistics,  
NC State University

:::: {.column width='33%'}

```{r, out.width = '100px'}
knitr::include_graphics('mtrosset.jpg')
```

Michael Trosset,  
Professor of Statistics,  
Indiana University

@@

Screenshot taken by User:DarwinPeacock - Screenshot of free software GUESS, CC BY 3.0, https://commons.wikimedia.org/w/index.php?curid=6057981

@@

\vspace*{1\baselineskip}

@@

## Community Detection in Block Models

Likelihood

$$L = \prod_{i<j} \prod_{k, l}^K 
\big(p_{k, l, i, j}^{A_{ij}} 
(1 - p_{k, l, i, j})^{1 - A_{ij}} \big)^{z_{ik} z_{jl}}$$

* ML method for community detection: $\hat{\vec{z}} = \arg\max_{\vec{z}} L$

* NP-complete
  * Expectation-Maximization
  * Bayesian methods
  * **Spectral methods**

# Generalized Random Dot Product Graphs

@@

Want to estimate $X$ given $A$.

@@

Rubin-Delanchy et al., 2020:  
$$\max_i \|\hat{X}_i - Q_n X_i \| \stackrel{a.s.}{\to} 0$$

@@

*Sparse Subspace Clustering* (Elhamifar and Vidal, 2009):  
Let the rows of $X \in \mathbb{R}^{n \times d}$ are vectors that lie on or near a small number of low-dimensional subspaces. SSC constructs a graph from $X$ as follows:

1. For each $i \in [n]$, solve the optimization problem  
$c_i = \arg\min_{c} \|c\|_1$ subject to $x_i = X c$ and $c^{(i)} = 0$.
2. Construct a similarity graph using $B = |C| + |C^\top|$ as its affinity matrix, where $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.

If $X$ obeys the *Subspace Detection Property*, then it produces a similarity graph $B$ such that $B_{ij} = 0$ $\forall z_i \neq z_j$.

## Sparse Subspace Clustering for Community Detection

@@

$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } A_{\cdot, i} = A c
\text{ and } c^{(i)} = 0$$

@@

$$c_i = \arg\min_{c} \|c\|_1 \text{ subject to } 
\hat{x}_i = \hat{X} c \text{ and } c^{(i)} = 0$$
$$A \approx \hat{X} I_{p, q} \hat{X}^\top$$

@@

@@

**Theorem** (KTT): 
If $P = V \Lambda V^\top$ is the spectral decomposition of $P$ for the PABM and 
$V$ has rows $v_i^\top$, then $v_i^\top v_j = 0$ $\forall z_i \neq z_j$.

**Theorem** (KTT): 
If $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$ is the spectral decomposition 
of $A$ for the PABM using the $K (K+1) / 2$ most positive and $K (K-1) / 2$ 
most negative eigenvalues, and we denote $\hat{v}_i^\top$ as the rows of 
$\hat{V}$, then

$$\max_{i, j : z_i \neq z_j} \hat{v}_i^\top \hat{v}_j = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)$$

Orthogonal Spectral Clustering (KTT):

1. Let $V$ be the eigenvectors of $A$ corresponding to the $K (K+1)/2$ most 
positive and $K (K-1) / 2$ most negative eigenvalues.
2. Compute $B = |n V V^\top|$ applying $|\cdot|$ entry-wise.
3. Construct graph $\hat{G}$ using $B$ as its similarity matrix.
4. Partition $\hat{G}$ into $K$ disconnected subgraphs.

## Simulation Study

Simulation setup:

1. $Z_1, ..., Z_n \stackrel{iid}{\sim} Categorical(1 / K, ..., 1 / K)$
2. $\lambda_{ik} \stackrel{iid}{\sim} Beta(a_{ik}, b_{ik})$  
$a_{ik} = \begin{cases} 2 & z_i = k \\ 1 & z_i \neq k \end{cases}$, 
$b_{ik} = \begin{cases} 1 & z_i = k \\ 2 & z_i \neq k \end{cases}$
3. $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
4. $A \sim BernoulliGraph(P)$

@@

## Conclusion

1. The PABM is a recently developed flexible block model that can be used to describe many graphs with community structure.

2. Likelihood maximization is NP-complete for block models, so we need to take a different approach to community detection.

3. The GRDPG, which describes all block models, motivates a spectral approach to statistical inference on graphs.

4. Under the GRDPG framework, the PABM is induced by a latent configuration in $\mathbb{R}^{K^2}$ consisting of $K$ $K$-dimensional subspaces that are all orthogonal to each other.

5. The latent configuration of the PABM under the GRDPG framework leads to two methods for community detection, both of which have nice theoretical asymptotic properties.

## Thank you

Code and draft of the paper available at  
https://github.com/johneverettkoo/pabm-grdpg

R package coming soon at https://github.com/johneverettkoo/pabm
